Model Finding for Recursive Functions in SMT\thanks{ This research is partially supported by the Inria technological development action ``Contre-exemples utilisables par Isabelle et Coq'' (CUIC). }

### 2016-01-25 ijcar2016/body.tex
well founded and a bit more

\newcommand\closurefmla{
  \ensuremath{\psi \models_T \bigwedge\nolimits_{\,i=1}^{n} \{ \varphi_i[ \vec t / \vec x ] \mid \con{f}_i( \vec t ) \in \terms( \psi ) \}} }

\begin{definition}\rm
Given a set of function definitions 
$\DDD = \{ \forallf{\con{f}_1} \vec x.\; \varphi_1, \ldots, \forallf{\con{f}_n} \vec x.\; \varphi_n \}$, 
a ground formula $\psi$ 
%\ct{I suppose $\psi$ is meant to be ground.} yes --ajr
is \emph{closed under function expansion with respect to $\DDD$} if
\begin{conf}$\closurefmla$. \end{conf}%
\begin{rep}\[\closurefmla\]\end{rep}%
The set $\DDD$ is \emph{admissible} if for every $T$-satisfiable formula 
$\psi$ closed under function expansion with respect to $\DDD$,
the formula
$\psi \wedge \bigwedge \Delta$ is also $T$-satisfiable.
%\ct{I}
\end{definition}

Admissibility is a semantic criterion that must be satisfied for each function
definition before applying the translation described in
Section~\ref{sec:encoding}. It is interesting to connect it to the
standard notion of \emph{well-founded} function definitions, often called
\emph{terminating} definitions in a slight abuse of terminology. In such
definitions, all recursive calls are decreasing with respect to a well-founded
relation, which must be supplied by the user or inferred automatically
using a termination prover. This ensures that the function is uniquely defined
at all points.

First-order logic has no built-in notion of computation or termination. To ensure
that a function specification is well-founded, it is sufficient to require that
the function would terminate when seen as a functional program, under \emph{some}
evaluation order. For example, the definition
$\forall x : {\typeint}.\;\, \lite\bigl(
x \leq 0,\allowbreak\; \con{p}( x ) \teq 1,\allowbreak\; \con{p}( x ) \teq 2 * \con{p}( x - 1 ) \bigr)$,
where $T$ is integer arithmetic extended with the uninterpreted symbol 
$\con p:\typeint \rightarrow \typeint$, can be shown well-founded under a strategy that
evaluates the condition of an $\lite$ before evaluating the relevant branch,
ignoring the other branch. \begin{rep}Logically, such dependencies can be captured by
congruence rules. \end{rep}Krauss developed these ideas in the more general context of
higher-order logic \cite[Section 2]{krauss-2009-phd}.

@@

\begin{rep}
Another useful fragment of function definitions is the class of
\emph{productive} corecursive functions. Corecursive functions are functions to
a coalgebraic datatype. These functions can be non-well-founded, without
their being inconsistent. Productive corecursive functions are functions that
progressively reveal parts of their potentially infinite
output \cite{turner-1995,mcbride-productive}.
For example, given a type of infinite streams constructed by
$\con{scons} : \ty{int} \times \ty{stream} \to \ty{stream}$,
the definition
$\forallf{\const{e}} x.\;\, \const{e}(x) \teq \con{scons}(x,\; \const{e}(x + 1))$
falls within this fragment: Each call to $\const{e}$ produces one
constructor before entering the nested call. Like terminating recursion,
productive corecursion totally specify the functions it defines.
It is even possible to mix recursion and corecursion in the same function
\cite{blanchette-et-al-2015-fouco}. Theorem~\ref{thm:adm} can be extended to
cover such specifications, based on the observation that unfolding
a corecursive definition infinitely computes a unique infinite object.

@@

\begin{description}
%{{{
  \item[Type inference] infers types and checks definitions;
  \item[Type skolemization]
    replaces $\exists \alpha. p[\alpha]$
      with $p[\tau]$ where $\tau$ is a fresh type;
  \item[Monomorphization]
    specializes polymorphic functions on their type arguments;
  \item[Elimination of Equations]
    translates multiple-equations definitions of functions into
      a single nested pattern matching;
  \item[Polarization]
    specializes some predicates into a version used in positive position
      and one used in negative position;
  \item[Unrolling]
    adds a decreasing argument to non-well-founded predicates;
  \item[Skolemization]
    introduces Skolem symbols for term variables;
  \item[Elimination of Predicates]
    changes a multi-clauses (co)inductive predicate definition
      into a single recursive function;
  \item[Recursion Elimination]
    performs the encoding from Section~\ref{sec:encoding};
  \item[Elimination of Match]
    transforms pattern-matching expressions into tests and selectors;
  \item[CVC4 Invocation] calls CVC4 to obtain a model.
%}}}
\end{description}

>>>>

\begin{definition}\rm
Given a set of function definitions 
$\DDD = \{ \forallf{\con{f}_1} \vec x.\; \varphi_1, \ldots, \forallf{\con{f}_n} \vec x.\; \varphi_n \}$, 
a ground formula $\psi$ 
%\ct{I suppose $\psi$ is meant to be ground.} yes --ajr
is \emph{closed under function expansion with respect to $\DDD$} if
\[\psi \models_T \smash{\bigwedge\nolimits_{\,i=1}^{n} \{ \varphi_i[ \vec t / \vec x ] \mid \con{f}_i( \vec t ) \in \terms( \psi ) \}}\]
The set $\DDD$ is \emph{admissible} if for every $T$-satisfiable formula 
$\psi$ closed under function expansion with respect to $\DDD$,
the formula
$\psi \wedge \bigwedge \Delta$ is also $T$-satisfiable.
%\ct{I}
\end{definition}

Admissibility is a semantic criterion that must be satisfied for each function
definition before applying the translation described in
Section~\ref{sec:encoding}. It is useful to connect it to the
standard notion of \emph{well-founded} function definitions, often called
\emph{terminating} definitions in a slight abuse of terminology. In such
definitions, all recursive calls are decreasing with respect to a well-founded
relation, which must be supplied by the user or inferred automatically
using a termination prover. This ensures that the function is uniquely defined
at all points.

First-order logic has no built-in notion of computation or termination. To ensure
that a function specification is well founded, it is sufficient to require that
the function would terminate when seen as a functional program, under \emph{some}
evaluation order. For example, the definition
$\forall x : {\typeint}.\;\, \lite\bigl(
x \leq 0,\allowbreak\; \con{p}( x ) \teq 1,\allowbreak\; \con{p}( x ) \teq 2 * \con{p}( x - 1 ) \bigr)$,
where $T$ is integer arithmetic extended with the uninterpreted symbol 
$\con p:\typeint \rightarrow \typeint$, can be shown well founded under a strategy that
evaluates the condition of an $\lite$ before evaluating the relevant branch,
ignoring the other branch. \begin{rep}Logically, such dependencies can be captured by
congruence rules. \end{rep}Krauss developed these ideas in the more general context of
higher-order logic \cite[Section 2]{krauss-2009-phd}.

@@

\begin{rep}
Another useful fragment of function definitions is the class of
\emph{productive} corecursive functions. Corecursive functions are functions to
a coalgebraic datatype. These functions can be ill founded, without
their being inconsistent. Productive corecursive functions are functions that
progressively reveal parts of their potentially infinite
output \cite{turner-1995,mcbride-productive}.
For example, given a type of infinite streams constructed by
$\con{scons} : \ty{int} \times \ty{stream} \to \ty{stream}$,
the definition
$\forallf{\const{e}} x.\;\, \const{e}(x) \teq \con{scons}(x,\; \const{e}(x + 1))$
falls within this fragment: Each call to $\const{e}$ produces one
constructor before entering the nested call. Like terminating recursion,
productive corecursion totally specify the functions it defines.
It is even possible to mix recursion and corecursion in the same function
\cite{blanchette-et-al-2015-fouco}. Theorem~\ref{thm:adm} can be extended to
cover such specifications, based on the observation that unfolding
a corecursive definition infinitely computes a unique infinite object.

@@

\begin{description}
%{{{
  \item[Type inference] infers types and checks definitions;
  \item[Type skolemization]
    replaces $\exists \alpha. p[\alpha]$
      with $p[\tau]$ where $\tau$ is a fresh type;
  \item[Monomorphization]
    specializes polymorphic functions on their type arguments;
  \item[Elimination of Equations]
    translates multiple-equations definitions of functions into
      a single nested pattern matching;
  \item[Polarization]
    specializes some predicates into a version used in positive position
      and one used in negative position;
  \item[Unrolling]
    adds a decreasing argument to possibly ill-founded predicates;
  \item[Skolemization]
    introduces Skolem symbols for term variables;
  \item[Elimination of Predicates]
    changes a multi-clauses (co)inductive predicate definition
      into a single recursive function;
  \item[Recursion Elimination]
    performs the encoding from Section~\ref{sec:encoding};
  \item[Elimination of Match]
    transforms pattern-matching expressions into tests and selectors;
  \item[CVC4 Invocation] calls CVC4 to obtain a model.
%}}}
\end{description}
